On the analytic structure of the linear response susceptibility in open systems

### 2018-05-02 AnalyticStructure_main.tex
update

Linear response can be effectively applied to the analysis of the electronic excitations in molecular systems. In this framework one consider a system described by the Hamiltonian 
$\hnot$ and assume that the ground-state density $\dmnot$ has been provided. The linear response formalism allows us to evaluate the modification of the expectation value of a generic 
observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed through the evaluation of the 
\emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;.
\ee
% where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
% under inspection. The modification of the ground-state density induced by the perturbation is expressed by the \emph{response density operator} $\dm'(\omega)$ that, in the linear 
% response regime, is a linear function of the perturbing field $\op\Phi(\omega)$. For this reason it is possible to define the \emph{susceptibility} superoperator, as follows:
% \be\lb{SusceptibilityDef1}
% \dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
% \ee
% Susceptibility describes an intrinsic feature of the unperturbed system and determines its response to \emph{any} perturbations, in the linear response regime.
Here $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
while the \emph{response density operator} $\dm'(\omega)$ describes the modification of the ground-state density induced by the perturbation. In the contest of the LR the response 
density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008}),
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field acting on the system.

@@

The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
Liouvillian. Pursuing this procedure provides:
%The knowledge of the excitations spectrum allows us to expand the LR susceptibility in terms of excitations. Indeed we can make usage of the spectral decomposition of the resolvent
%of the Liouvillian. This procedure allows us to formally solve the Liouville equation \eqref{LiouvillianRhopomegaDef1} and we have 
\be\lb{RhopExcitationDef1}
\sop{{\cal R}}(\omega) \cdot   =
\left(\sum_E + \int\dd E   \right) \op E \,
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
\ee 
%where we have divided the two sectors in the sum. 
where we have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef1}. 
The elements of the first sector correspond to the \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

@@

\subsection{Discrete poles and spectral function. Physical meaning of the excitation spectrum} %{Comments. Physical meaning of the excitation spectrum}

The analysis discussed above has evidenced that there exists both localized and delocalized excitations. While the role of the former is quite unambigous 
We can express the fluctuation states in terms of the susceptibility functional
\be\lb{FluctuationStateExcitationDef1}
\ket{f_p(\omega)}   =
\left(\sum_E + \int\dd E   \right) 
%\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}}{\omega-E} \ket{\phi^E_p} \;. \nn
\frac{t_\Phi(E)}{\omega-E} \ket{\phi^E_p} \;,
\ee 
with $t_\Phi(E)=\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}$.
This relation clarifies the role of the excitations in the spectral decomposition. We know from the analysis of section \ref{FluctuationState} that fluctuation states are localized
up to $\omega=|\eps_h|$.

\section{Criterion for a reliable sampling of the excitation operators} %Applications

>>>>

Linear response can be effectively applied to the analysis of the electronic excitations in molecular systems. In this framework one consider a system described by the Hamiltonian 
$\hnot$ and assume that the ground-state density $\dmnot$ has been provided. The linear response formalism allows us to evaluate the modification of the expectation value of a generic 
observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed through the evaluation of the 
\emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;.
\ee
Here $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
while the \emph{response density operator} $\dm'(\omega)$ describes the modification of the ground-state density induced by the perturbation. In the contest of the LR the response 
density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008}),
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field acting on the system.

@@

The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
Liouvillian. Pursuing this procedure provides:
%The knowledge of the excitations spectrum allows us to expand the LR susceptibility in terms of excitations. Indeed we can make usage of the spectral decomposition of the resolvent
%of the Liouvillian. This procedure allows us to formally solve the Liouville equation \eqref{LiouvillianRhopomegaDef1} and we have 
\be\lb{RhopExcitationDef1}
\sop{{\cal R}}(\omega) \cdot   =
\left(\sum_E + \int\dd E   \right) \op E \,
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
\ee 
where we have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef1}. The elements of the first sector 
correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

Lastly, we establish a link between the fluctuation states and excitation operators based description of the LR. Indeed, thanks to the defining equation \eqref{SusceptibilityDef1} we can 
easily express the excitations expansion of fluctuation states as follows 
\be\lb{FluctuationStateExcitationDef1}
\ket{f_p(\omega)}   =
\left(\sum_E + \int\dd E   \right) 
%\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}}{\omega-E} \ket{\phi^E_p} \;. \nn
\frac{t_\Phi(E)}{\omega-E} \ket{\phi^E_p} \;,
\ee 
with $t_\Phi(E)=\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}$. 
%This formula shows that, for each value of $\omega$, both the sectors of the excitations spectrum contribute to the fluctuation state. 
Thanks to the analysis presented in section \ref{FluctuationState} we know that fluctuation states possess a localized behavior up to $\omega=|\eps_h|$, so we conclude that 
below this energy threshold the continuum sector of the excitation spectrum is able to resum in a localized way.

@@

%\subsection{Discrete poles and spectral function. Physical meaning of the excitation spectrum} %{Comments. Physical meaning of the excitation spectrum}

% The analysis discussed above has evidenced that there exists both localized and delocalized excitations. While the role of the former is quite unambigous 
% We can express the 
% \be\lb{FluctuationStateExcitationDef1}
% \ket{f_p(\omega)}   =
% \left(\sum_E + \int\dd E   \right) 
% %\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}}{\omega-E} \ket{\phi^E_p} \;. \nn
% \frac{t_\Phi(E)}{\omega-E} \ket{\phi^E_p} \;,
% \ee 
% with $t_\Phi(E)=\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}$.
% This relation clarifies the role of the excitations in the spectral decomposition. We know from the analysis of section \ref{FluctuationState} that fluctuation states are localized
% up to $\omega=|\eps_h|$.

\section{Quality indicator for the numerical convergence of the LR}   %{Criterion for a reliable sampling of the excitation operators}

From the theoretical analysis of the previous section we can made two statements:
\begin{itemize}
 \item Fluctuation states, which constitute the building blocks of the response density, have a bound state-like character up to $\omega=|\eps_h|$ and behave as unbound delocalized
 wave functions for higher values of $\omega$. 
 \item The excitation landscape is composed of two sectors with distinct properties. One made of a discrete set of localized excitations 
\end{itemize}
